Lessons learned: creating tutorials to teach agent-based modelling to archaeologists

### 2025-09-01 paper.Rmd
Minor correction and generate png-files for paper graphs

---
title: "Lessons learned: creating tutorials to teach agent-based modelling to archaeologists"
authors:
  - name: Ronald M. Visser
    orcid_id: 0000-0001-6966-1729
    roles: [writing, editing, analysis, visualization]
    affiliation: Saxion University of Applied Sciences
  - name: Karsten Lambers 
    roles: [writing, editing]
    orcid_id: 0000-0001-6432-0925
    affiliation: Leiden University
  - name: Kenneth Aitchison 
    roles: [writing, editing, software, project administration]
    orcid_id: 0000-0001-6594-1408
    affiliation: Landward Research Teoranta
  - name: Annemarie Jutte
    orcid_id: 0009-0004-9322-0674
    roles: [writing, editing, software]
    affiliation: Saxion University of Applied Sciences
  - name: Doug Rocks-Macqueen 
    orcid_id: 0000-0001-9021-0260
    roles: [writing, editing, software]
    affiliation: Landward Research Teoranta
  - name: Laura van der Knaap 
    roles: [writing, editing, software]
    orcid_id: 0000
    affiliation: Leiden University
  - name: Iza Romanowska 
    orcid_id: 0000-0002-9487-2111
    roles: [writing, editing]
    affiliation: Aarhus University
  - name: Tom Brughmans
    roles: [writing, editing]
    orcid_id: 0000-0002-1589-7768
    affiliation: Aarhus University
  - name: Jeroen Linssen
    orcid_id: 0000-0002-2626-1837
    roles: [writing, editing]
    affiliation: Saxion University of Applied Sciences
  - name: Adam Pažout
    roles: [writing, editing]
    orcid_id: 0000-0001-7745-5634
    affiliation: Aarhus University
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  distill::distill_article:
    number_sections: yes
    fig_caption: yes
  bookdown::pdf_document2:
    number_sections: yes
    keep_tex: true
    fig_caption: yes
  bookdown::word_document2:
    number_sections: yes
    fig_caption: yes
bibliography: references.bib
csl: journal-of-computer-applications-in-archaeology.csl
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding,
  output_format = "all") })
---

@@

**Abstract**

Agent-based modelling (ABM) is one of the most widely applied computational modelling research techniques in archaeology, because it is well suited to the questions asked by archaeologists. However, the access to vocational training for ABM is limited. In this paper we present the Open Educational Resources (OER) that we produced in an international cooperation. For this purpose, we created online and open tutorials for ABM for archaeologists. In addition, how-to guides and support material were created, both for teachers and students. The materials adhere to international (European) standards in relation to education and the Digital Skills Passport for archaeology. The teaching materials have been well tested using a diverse group of archaeologists during various international conferences and meetings. It was shown that the knowledge and skills of the participants in relation to ABM improved. In addition, the demand for the teaching materials and the workshops was high, ensuring that the tutorials on ABM will be used in the future.

@@

```{r nationality, fig.cap="The nationality of the participants that filled in the survey.", fig.height=5, echo=FALSE}
pre_surveys %>% 
  group_by(`What.is.you.nationality....Nationality...Text`) %>% 
  mutate(count_nationality = n()) %>%
  ggplot(aes(x=reorder(`What.is.you.nationality....Nationality...Text`, count_nationality))) + geom_bar() +  coord_flip() + xlab("Nationality")

@@

```{r gender-age, fig.cap="The gender and age distribution of the respondents for each workshop.", fig.width=11, echo=FALSE}
ggplot(pre_surveys, aes(x = `What.is.your.age.`, fill = `What.is.your.gender....Selected.Choice`)) +
  geom_bar() + labs(x= "Age", y="Count", fill = "Gender") + facet_grid(~Workshop, labeller = event_names) +   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

@@

```{r, computer-skills, fig.cap="The computer skills of the respondents.", out.width="50%", echo=FALSE}
pre_surveys %>%
  select(`Which.computer.skills.do.you.have....Selected.Choice`) %>%
  separate_rows(`Which.computer.skills.do.you.have....Selected.Choice`, sep = ',\\s*') %>%
  filter(`Which.computer.skills.do.you.have....Selected.Choice`!="") %>%
  group_by(`Which.computer.skills.do.you.have....Selected.Choice`) %>%
    mutate(count_skills = n()) %>%
  ggplot(aes(x=reorder(`Which.computer.skills.do.you.have....Selected.Choice`, count_skills))) + geom_bar() +
    xlab("Computer skills") + coord_flip()
```

```{r, abm-knowledge, fig.cap="Respondents knowledge of ABM per event (top), and knowledge of the various softwares for ABM (bottom), before participating in the workshops.", echo=FALSE}
knowledge_plot <- pre_surveys %>%
  filter(`Do.you.know.what.Agent.Based.Modeling.is.`!="Not shared") %>%
  select(Do.you.know.what.Agent.Based.Modeling.is., Workshop) %>%
  group_by(`Do.you.know.what.Agent.Based.Modeling.is.`, Workshop) %>%
  mutate(count_response = n()) %>%
  distinct() %>%
  ggplot(aes(fill = `Do.you.know.what.Agent.Based.Modeling.is.`, x=Workshop, y = count_response)) +
  geom_bar(na.rm = TRUE, position="fill", stat="identity") +
  geom_text(aes(label=count_response), position = position_fill(vjust = 0.5), size = 2.5, colour = "grey41") + scale_x_discrete(labels = c("CAA", "EAA", "CAA-DE-NLFl", "Reuvensdagen", "CAA_UK", "Leiden", "Aarhus")) +
  xlab("") + scale_fill_discrete(name = "ABM known") + ylab("Percentage") +  scale_y_continuous(labels = scales::label_percent(scale = 100, prefix = "", suffix = "")) + coord_flip()
software_plot <- pre_surveys %>%
  select(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`) %>%
  filter(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`!="") %>%
  separate_rows(Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice, sep = ',\\s*') %>%
  filter(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`!="None") %>%
  group_by(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`) %>%
    mutate(count_software = n()) %>%
  ggplot(aes(x=reorder(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`, count_software))) + geom_bar() + scale_x_discrete(labels = ) + xlab("ABM Software known") + coord_flip()
ggarrange(knowledge_plot, software_plot, nrow = 2, ncol = 1)
```

@@

```{r, available-theory, fig.cap="Respondents opinion on the quality of theory on ABM faceted out by event.", fig.width=11, echo=FALSE}
pre_surveys %>%
  select(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`, Workshop) %>%
  filter(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`!="") %>%
  mutate(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.` =
           fct_relevel(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`,
                       "No opinion", "Limited", "Sufficient", "More than sufficient", "Very good")) %>%
  ggplot(aes(x = `How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`)) +
  geom_bar() + xlab("Quality of theory on ABM") + coord_flip() +
  facet_grid(~Workshop, labeller = event_names) + 
  scale_x_discrete(drop=FALSE)
```

@@

```{r, rating-workshop, fig.cap="Respondents rating of the workshop in general faceted for each event.", fig.width=11, echo=FALSE}
post_surveys %>%
  mutate(How.would.you.rate.the.workshop.in.general. = factor(How.would.you.rate.the.workshop.in.general., factor_levels)) %>%
  ggplot(aes(x=How.would.you.rate.the.workshop.in.general.)) + geom_bar() +
  xlab("Rating of the workshop") + facet_grid(~Workshop, labeller = event_names) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  scale_x_discrete(drop=FALSE)

@@

```{r, rating-teaching, fig.cap="Respondents rating of the teaching material (top) and teachers (bottom) in general faceted for each event.", fig.width=11, fig.height=8, echo=FALSE}
rate_material_plot <- post_surveys %>%
  mutate(How.would.you.rate.the.teaching.material. = factor(How.would.you.rate.the.teaching.material., factor_levels)) %>%
  ggplot(aes(x=How.would.you.rate.the.teaching.material.)) + geom_bar() +
  xlab("Rating of the teaching material") + facet_grid(~Workshop, labeller = event_names) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  scale_x_discrete(drop=FALSE)
rate_teacher_plot <- post_surveys %>%
  mutate(How.would.you.rate.the.trainers.teachers. = factor(How.would.you.rate.the.trainers.teachers., factor_levels)) %>%
  ggplot(aes(x=How.would.you.rate.the.trainers.teachers.)) + geom_bar() +
  xlab("Rating of the teacher") + facet_grid(~Workshop, labeller = event_names) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  scale_x_discrete(drop=FALSE)
ggarrange(rate_material_plot, rate_teacher_plot, nrow = 2, ncol = 1)
```

>>>>

---
title: "Lessons learned: creating tutorials to teach agent-based modelling to archaeologists"
authors:
  - name: Ronald M. Visser
    orcid_id: 0000-0001-6966-1729
    roles: [writing, editing, analysis, visualization]
    affiliation: Saxion University of Applied Sciences
  - name: Karsten Lambers 
    roles: [writing, editing]
    orcid_id: 0000-0001-6432-0925
    affiliation: Leiden University
  - name: Kenneth Aitchison 
    roles: [writing, editing, software, project administration]
    orcid_id: 0000-0001-6594-1408
    affiliation: Landward Research Teoranta
  - name: Annemarie Jutte
    orcid_id: 0009-0004-9322-0674
    roles: [writing, editing, software]
    affiliation: Saxion University of Applied Sciences
  - name: Doug Rocks-Macqueen 
    orcid_id: 0000-0001-9021-0260
    roles: [writing, editing, software]
    affiliation: Landward Research Teoranta
  - name: Laura van der Knaap 
    roles: [writing, editing, software]
    orcid_id: 0000
    affiliation: Leiden University
  - name: Iza Romanowska 
    orcid_id: 0000-0002-9487-2111
    roles: [writing, editing]
    affiliation: Aarhus University
  - name: Tom Brughmans
    roles: [writing, editing]
    orcid_id: 0000-0002-1589-7768
    affiliation: Aarhus University
  - name: Jeroen Linssen
    orcid_id: 0000-0002-2626-1837
    roles: [writing, editing]
    affiliation: Saxion University of Applied Sciences
  - name: Adam Pažout
    roles: [writing, editing]
    orcid_id: 0000-0001-7745-5634
    affiliation: Aarhus University
date: '`r format(Sys.Date(), "%B %d, %Y")`'
keywords: ["agent-based modelling", "education", "open educational resources", "archaeological skills", "computational skills", "simulation", "teaching"]
abstract: "Agent-based modelling (ABM) is one of the most widely applied computational modelling research techniques in archaeology, because it is well suited to the questions asked by archaeologists. However, the access to vocational training for ABM is limited. In this paper we present the Open Educational Resources (OER) that we produced in an international cooperation. For this purpose, we created online and open tutorials for ABM for archaeologists. In addition, how-to guides and support material were created, both for teachers and students. The materials adhere to international (European) standards in relation to education and the Digital Skills Passport for archaeology. The teaching materials have been well tested using a diverse group of archaeologists during various international conferences and meetings. It was shown that the knowledge and skills of the participants in relation to ABM improved. In addition, the demand for the teaching materials and the workshops was high, ensuring that the tutorials on ABM will be used in the future."
output:
  distill::distill_article:
    number_sections: yes
    fig_caption: yes
    keep_md: yes
  bookdown::pdf_document2:
    number_sections: yes
    keep_tex: true
    fig_caption: yes
  bookdown::word_document2:
    number_sections: yes
    fig_caption: yes
bibliography: references.bib
csl: journal-of-computer-applications-in-archaeology.csl
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding,
  output_format = "all") })
---

@@

@@

```{r nationality, dpi=600, fig.cap="The nationality of the participants that filled in the survey.", fig.height=5, echo=FALSE}
pre_surveys %>% 
  group_by(`What.is.you.nationality....Nationality...Text`) %>% 
  mutate(count_nationality = n()) %>%
  ggplot(aes(x=reorder(`What.is.you.nationality....Nationality...Text`, count_nationality))) + geom_bar() +  coord_flip() + xlab("Nationality")

@@

```{r gender-age, dpi=600, fig.cap="The gender and age distribution of the respondents for each workshop.", fig.width=11, echo=FALSE}
ggplot(pre_surveys, aes(x = `What.is.your.age.`, fill = `What.is.your.gender....Selected.Choice`)) +
  geom_bar() + labs(x= "Age", y="Count", fill = "Gender") + facet_grid(~Workshop, labeller = event_names) +   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

@@

```{r, computer-skills, dpi=600, fig.cap="The computer skills of the respondents.", out.width="50%", echo=FALSE}
pre_surveys %>%
  select(`Which.computer.skills.do.you.have....Selected.Choice`) %>%
  separate_rows(`Which.computer.skills.do.you.have....Selected.Choice`, sep = ',\\s*') %>%
  filter(`Which.computer.skills.do.you.have....Selected.Choice`!="") %>%
  group_by(`Which.computer.skills.do.you.have....Selected.Choice`) %>%
    mutate(count_skills = n()) %>%
  ggplot(aes(x=reorder(`Which.computer.skills.do.you.have....Selected.Choice`, count_skills))) + geom_bar() +
    xlab("Computer skills") + coord_flip()
```

```{r, abm-knowledge, dpi=600, fig.cap="Respondents knowledge of ABM per event (top), and knowledge of the various softwares for ABM (bottom), before participating in the workshops.", echo=FALSE}
knowledge_plot <- pre_surveys %>%
  filter(`Do.you.know.what.Agent.Based.Modeling.is.`!="Not shared") %>%
  select(Do.you.know.what.Agent.Based.Modeling.is., Workshop) %>%
  group_by(`Do.you.know.what.Agent.Based.Modeling.is.`, Workshop) %>%
  mutate(count_response = n()) %>%
  distinct() %>%
  ggplot(aes(fill = `Do.you.know.what.Agent.Based.Modeling.is.`, x=Workshop, y = count_response)) +
  geom_bar(na.rm = TRUE, position="fill", stat="identity") +
  geom_text(aes(label=count_response), position = position_fill(vjust = 0.5), size = 2.5, colour = "grey41") + scale_x_discrete(labels = c("CAA", "EAA", "CAA-DE-NLFl", "Reuvensdagen", "CAA_UK", "Leiden", "Aarhus")) +
  xlab("") + scale_fill_discrete(name = "ABM known") + ylab("Percentage") +  scale_y_continuous(labels = scales::label_percent(scale = 100, prefix = "", suffix = "")) + coord_flip()
software_plot <- pre_surveys %>%
  select(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`) %>%
  filter(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`!="") %>%
  separate_rows(Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice, sep = ',\\s*') %>%
  filter(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`!="None") %>%
  group_by(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`) %>%
    mutate(count_software = n()) %>%
  ggplot(aes(x=reorder(`Which.software.do.you.know.for.Agent.Based.Modeling....Selected.Choice`, count_software))) + geom_bar() + scale_x_discrete(labels = ) + xlab("ABM Software known") + coord_flip()
ggarrange(knowledge_plot, software_plot, nrow = 2, ncol = 1)
```

@@

```{r, available-theory, dpi=600, fig.cap="Respondents opinion on the quality of theory on ABM faceted out by event.", fig.width=11, echo=FALSE}
pre_surveys %>%
  select(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`, Workshop) %>%
  filter(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`!="") %>%
  mutate(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.` =
           fct_relevel(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`,
                       "No opinion", "Limited", "Sufficient", "More than sufficient", "Very good")) %>%
  ggplot(aes(x = `How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`)) +
  geom_bar() + xlab("Quality of theory on ABM") + coord_flip() +
  facet_grid(~Workshop, labeller = event_names) + 
  scale_x_discrete(drop=FALSE)
```

@@

```{r, rating-workshop, dpi=600, fig.cap="Respondents rating of the workshop in general faceted for each event.", fig.width=11, echo=FALSE}
post_surveys %>%
  mutate(How.would.you.rate.the.workshop.in.general. = factor(How.would.you.rate.the.workshop.in.general., factor_levels)) %>%
  ggplot(aes(x=How.would.you.rate.the.workshop.in.general.)) + geom_bar() +
  xlab("Rating of the workshop") + facet_grid(~Workshop, labeller = event_names) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  scale_x_discrete(drop=FALSE)

@@

```{r, rating-teaching, dpi=600, fig.cap="Respondents rating of the teaching material (top) and teachers (bottom) in general faceted for each event.", fig.width=11, fig.height=8, echo=FALSE}
rate_material_plot <- post_surveys %>%
  mutate(How.would.you.rate.the.teaching.material. = factor(How.would.you.rate.the.teaching.material., factor_levels)) %>%
  ggplot(aes(x=How.would.you.rate.the.teaching.material.)) + geom_bar() +
  xlab("Rating of the teaching material") + facet_grid(~Workshop, labeller = event_names) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  scale_x_discrete(drop=FALSE)
rate_teacher_plot <- post_surveys %>%
  mutate(How.would.you.rate.the.trainers.teachers. = factor(How.would.you.rate.the.trainers.teachers., factor_levels)) %>%
  ggplot(aes(x=How.would.you.rate.the.trainers.teachers.)) + geom_bar() +
  xlab("Rating of the teacher") + facet_grid(~Workshop, labeller = event_names) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  scale_x_discrete(drop=FALSE)
ggarrange(rate_material_plot, rate_teacher_plot, nrow = 2, ncol = 1)
```

### 2025-09-01 paper.tex
Minor correction and generate png-files for paper graphs

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Lessons learned: creating tutorials to teach agent-based modelling to archaeologists},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Lessons learned: creating tutorials to teach agent-based modelling to archaeologists}
\author{}
\date{\vspace{-2.5em}August 26, 2025}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\textbf{Abstract}

Agent-based modelling (ABM) is one of the most widely applied computational modelling research techniques in archaeology, because it is well suited to the questions asked by archaeologists. However, the access to vocational training for ABM is limited. In this paper we present the Open Educational Resources (OER) that we produced in an international cooperation. For this purpose, we created online and open tutorials for ABM for archaeologists. In addition, how-to guides and support material were created, both for teachers and students. The materials adhere to international (European) standards in relation to education and the Digital Skills Passport for archaeology. The teaching materials have been well tested using a diverse group of archaeologists during various international conferences and meetings. It was shown that the knowledge and skills of the participants in relation to ABM improved. In addition, the demand for the teaching materials and the workshops was high, ensuring that the tutorials on ABM will be used in the future.

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Lessons learned: creating tutorials to teach agent-based modelling to archaeologists},
  pdfkeywords={agent-based modelling, education, open educational resources, archaeological skills, computational skills, simulation, teaching},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Lessons learned: creating tutorials to teach agent-based modelling to archaeologists}
\author{}
\date{\vspace{-2.5em}September 01, 2025}

\begin{document}
\maketitle
\begin{abstract}
Agent-based modelling (ABM) is one of the most widely applied computational modelling research techniques in archaeology, because it is well suited to the questions asked by archaeologists. However, the access to vocational training for ABM is limited. In this paper we present the Open Educational Resources (OER) that we produced in an international cooperation. For this purpose, we created online and open tutorials for ABM for archaeologists. In addition, how-to guides and support material were created, both for teachers and students. The materials adhere to international (European) standards in relation to education and the Digital Skills Passport for archaeology. The teaching materials have been well tested using a diverse group of archaeologists during various international conferences and meetings. It was shown that the knowledge and skills of the participants in relation to ABM improved. In addition, the demand for the teaching materials and the workshops was high, ensuring that the tutorials on ABM will be used in the future.
\end{abstract}

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}
